class: center, middle, inverse

# [`dplyr`](https://dplyr.tidyverse.org/)

.footnote[
https://dplyr.tidyverse.org/
]

???

* dplyr is a grammar of data manipulation *on tidy data*
* (Relatively) consistent
* Provide few 'verbs' to do most things --- one-way philosophy
* Fast, not not built for speed. `data.table` might be better here.

From the docs:pp

> * Identify the most important data manipulation verbs and make them easy to use from R.
> * Provide blazing fast performance for in-memory data by writing key pieces in C++ (using Rcpp)
> * Use the same interface to work with data no matter where it's stored, whether in a data frame, a data table or database.

---
### dplyr overview

Core functionality:

* `select()` columns
* `filter()` rows
* `arrange()` / sort rows
* `mutate()` and `transmute()`: add new columns

Reduce/summarize (groups of) rows with:

* `summarise()`, `summarize()`

and 

* `group_by()`, `ungroup()`



---
### `mtcars`


```{r}
mtcars %>% print(n = 5, width = 60)
```

---
### Basic dplyr in action
```{r}
mtcars %>% 
  filter(cyl %>% between(4,6)) #<<
```

???

* `filter` has helper functions `between`, `near`, ``xor`

---
### Basic dplyr in action
```{r}
mtcars %>% 
  filter(cyl %>% between(4,6)) %>% 
  select(car, mpg:wt, -drat) # also supports -(colX:colY)  #<<
```

???

* `select` has helper functions `starts_with`, `ends_with`, `contains`, `matches`


---
### Basic dplyr in action
```{r}
mtcars %>% 
  filter(cyl %>% between(4,6)) %>% 
  select(car, mpg:wt, -drat) %>%
  mutate(wt = 0.45*wt, `hp/wt` = hp/wt) #<<
```

???

* `mutate` supports multiple new columns, created in order
    - mtcars %>% mutate(cyl_disp_ccm = 16.387064*disp/cyl, cyl_disp_L = cyl_disp_ccm/1000)
* `mutate` has helper functions `cumall`, `cumany`, `recode`, `case_when`, `percent_rank`
* `transmute` would just return the derived value

---
### Basic dplyr in action
```{r}
mtcars %>% 
  filter(cyl %>% between(4,6)) %>%
  select(car, mpg:wt, -drat) %>%
  mutate(wt = 0.45*wt, `hp/wt` = hp/wt) %>%
  arrange(desc(`hp/wt`)) #<<
```

---
### Basic dplyr in action
```{r}
mtcars %>% 
  filter(cyl %>% between(4,6)) %>% 
  select(car, mpg:wt, -drat) %>%
  mutate(wt = 0.45*wt, `hp/wt` = hp/wt) %>% 
  arrange(desc(`hp/wt`)) %>% 
  group_by(cyl) #<<
```

---
### Basic dplyr in action
```{r}
mtcars %>% 
  filter(cyl %>% between(4,6)) %>% 
  select(car, mpg:wt, -drat) %>%
  mutate(wt = 0.45*wt, `hp/wt` = hp/wt) %>%
  arrange(desc(`hp/wt`)) %>% 
  group_by(cyl) %>%  
  summarize(mean_hpwt = mean(`hp/wt`), tanh = tanh(mean_hpwt)) #<<
```

???

* Doing the above it "[" would be a nightmare in comparison
* Again lazy evaluation, in order is generally supported.
* 'Context-reference' (NSE as Janus talked about last time)

---
class: center, middle

# [`dbplyr`](https://dbplyr.tidyverse.org/)

.footnote[
https://dbplyr.tidyverse.org/
]

???

* `dbplyr` is a database back-end for `dplyr`
* Unlike "[", it abstracts away how the data is stored
* Converts `dplyr` code to SQL calls
* Expressions are lazily evaluated and `dplyr` pipes generate SQL, which is sent to the DB only when requested.
