class: center, middle, inverse

# [`purrr`](https://purrr.tidyverse.org/)

.footnote[
https://purrr.tidyverse.org/
]


---
class: middle
### Functional programming

* Abstracts away for-loops
* Consistent interface for working with vectors (incl. lists) and functions
* Alternative to the `apply` family 

???

* Code should be easier to read and reason about
* Downside, I think it gets quite complex quit easily (but I'm not used to reason that way)



---
### Motivation
```{r, echo = -1}
set.seed(1)
tbl <- tibble(a = rnorm(10, mean = 100), b = rnorm(10, mean = 100),
              c = rnorm(10, mean = 100), d = rnorm(10, mean = 100))
print(tbl)
```
Say we want to compute the standard deviation for each column.


---
### Motivation (cont)
```{r}
sd(tbl$a)
sd(tbl$b)
sd(tbl$c)
sd(tbl$d)
```

---
### Motivation (cont)
```{r}
out <- vector("numeric", ncol(tbl))
for (i in seq_along(out)) {
  out[i] <- sd(tbl[[i]]) #<<
}
out
```


---
### Motivation (cont)
```{r}
out2 <- sapply(tbl, sd)
out2
```

---
### Motivation (cont)
```{r}
out3 <- purrr::map(tbl, sd)  # basically identical with lapply
out3
```

---
### Motivation (cont)
```{r}
out4 <- purrr::map_dbl(tbl, sd)
out4
```

---
### Motivation (cont)
```{r}
out5 <- purrr::map_chr(tbl, sd)
out5
```


---
### Remember tibbles (and data.frames) are lists
```{r}
mtcars %>%
  split(.$cyl) %>% # from base R
  map(~ lm(mpg ~ wt, data = .)) %>%
  map(summary) %>%
  map_dbl("r.squared")
```



???

<!-- from https://purrr.tidyverse.org/ -->
* The first argument is always the data, so `purrr` works naturally with the pipe.
All purrr functions are type-stable. 
* They always return the advertised output type (map() returns lists; map_dbl() returns double vectors), or they throw an error.
* All map() functions either accept function, formulas (used for succinctly generating anonymous functions), a character vector (used to extract components by name), or a numeric vector (used to extract by position).



---
class: middle

### Multivariate map
```{r}
mu <- list(5, 10, -3)
sigma <- list(1, 3, 6)
map2(mu, sigma, rnorm, n = 4)
```

* `pmap` generalizes `map2` this to p-arguments (to avoid `map3`, `map4`, ...)



---
#### Invoking different functions

```{r}
f <- c("runif", "rnorm", "rpois")
param <- list(
  list(min = -1, max = 1), 
  list(sd = 5), 
  list(lambda = 10)
)
invoke_map(f, param, n = 5) 
```

---
#### Works well with `tibbles`
```{r}
sim <- tribble(
  ~f,      ~params,
  "runif", list(min = -1, max = 1),
  "rnorm", list(sd = 5),
  "rpois", list(lambda = 10)
)
print(sim)
sim %>% 
  mutate(sim = invoke_map(f, params, n = 10))
```



---
### `reduce` & `accumulate`
```{r}
vs <- list(
  c(1, 3, 5, 6, 10),
  c(1, 2, 3, 7, 8, 10),
  c(1, 2, 3, 4, 8, 9, 10)
)
vs %>% reduce(intersect)

x <- c(1,5,2,3)
x %>% accumulate(`-`)  # cummulative difference
```






