class: center, middle, inverse

# [`stringr`](https://stringr.tidyverse.org/)

.footnote[
https://stringr.tidyverse.org/
]


---
class: middle

### Consistent string manipulation functions
* `str_` prefixed functions

```{r}
ls("package:stringr")
```

???
  
* Consistent interface to string manipulation
* Of course 

---
  class: middle

```{r}
ls("package:stringr") %>% head(n = 18) %>% 
str_view_all("str_|ex")
```

???
  
* I want to highlight `str_view_all` which brings up a html_widget to show matches.
- Helpful when making a regex





---
class: center, middle, inverse

# [`forcats`](https://forcats.tidyverse.org/)

.footnote[
https://forcats.tidyverse.org/
]

???
  
>  The goal of the `forcats` package is to provide a suite of tools that solve common problems with factors, including changing the order of levels or the values. 


---
  
###  Consistent manipulating of factors
  
  
```{r}
ls("package:forcats")
```


---
```{r, }
fct_explicit_na( factor(c("A","B",NA)) )
fct_c( factor(c("A","B") ), factor(c("B","C")) )
fct_rev( factor(c("A","B")) )
fct_drop( factor(c("A","B", NA), levels = c("A","B","C")) )
fct_anon (factor(c("A","B",NA)) , prefix = "cat" )
```

???
  
* Some forcats examples


