class: center, middle, inverse

# [`tibble`](https://tibble.tidyverse.org/)

.footnote[
https://tibble.tidyverse.org/
]



---
### tibbles

```{r}
tbl <- tibble(x = 1:50, y = exp(x), w = y > 1, char = "AaRUG")
tbl
class(tbl)
```

???

Much like the data.frame but with all the annoying stuff taken away.
Creating tibbles are covered in syntactic sugar.
It extends the data.frame; but actually "dumber."
All tibbles are data.frames; but not all data.frames are tibbles.

They feature:
* Better printing
* Only recycles length 1 inputs.
* Evaluates its arguments lazily and in order.
* Never coerces inputs (i.e. strings stay as strings!).
* Never adds row.names.
* Never munges column names.
* Adds tbl_df class to output.
* Automatically adds column names.



---
### tibbles (cont.)

Subsetting via "[" does not "drop". Subsetting with "$" does.
```{r}
tbl[25, ]
print(tbl[, "y"], n = 3)
str(tbl$y)
```


---
### tibbles (cont.)

Also no partial matching!

```{r}
tbl$ch
data.frame(a = 1, char = "test")$ch
```



### tibbles do less and compain more

???

Hopefully this should lead to more expressive code and confront problems



