class: center, middle, inverse

# [`tidyr`](https://tidyr.tidyverse.org/)

.footnote[
https://tidyr.tidyverse.org/
]


---
class: center, middle

> The goal of `tidyr` is to help you create **tidy data**.

.footnote[
https://tidyr.tidyverse.org/
]



---
### Key tidyr functions

* `gather()`: *gathers* multiple columns into two key-value columns
  - i.e. wide to long format

* `spread()`: *spreads* two columns (key & value) into multiple columns
  - i.e. long to wide format

* `separate()`: pulls apart one `character` column into many (inverse of `unite()`)
  - `separate_rows` separate into extra rows
  
* `extract()`: similar, but uses regex to capture groups


???

* Alternative to `reshape` and `reshape2` packages.



---
### Other tidying
#### Handle missing values
* `drop_na()`  filters `NA`
* `fill()`  fills `NA` with most recent non-`NA` (from top and bottom)
* `replace_na()`
* `complete()`: Expand current tibble to make missing values explicit
* `expand()`: Creates a new tibble (like `expand.grid`)


---
### tidyr in action
```{r, echo = -1}
WHO_tuberculosis <- table5[-6, ]
WHO_tuberculosis #<<
```

.footnote[
http://www.who.int/tb/country/data/download/en/
`tidyr::table5`
]

---
### tidyr in action
```{r}
WHO_tuberculosis %>% 
  unite(col = "year", century, year, sep = "")   #<<
```

---
### tidyr in action
```{r}
WHO_tuberculosis %>% 
  unite(col = "year", century, year, sep = "") %>% 
  separate(rate, into = c("cases", "pop"))  #<<
```

---
### tidyr in action
```{r}
WHO_tuberculosis %>% 
  unite(col = "year", century, year, sep = "") %>% 
  separate(rate, into = c("cases", "pop")) %>% 
  gather(cases, pop, key = type, value = count) #<<
```

---
### tidyr in action
```{r}
WHO_tuberculosis %>% 
  unite(col = "year", century, year, sep = "") %>% 
  separate(rate, into = c("cases", "pop")) %>% 
  gather(cases, pop, key = type, value = count) %>% 
  complete(country, year, type) #<<
```

---
### tidyr in action
```{r}
WHO_tuberculosis %>% 
  unite(col = "year", century, year, sep = "") %>% 
  separate(rate, into = c("cases", "pop")) %>% 
  gather(cases, pop, key = type, value = count) %>% 
  complete(country, year, type) %>% 
  replace_na() #<<
```











