
class: center, middle
# [tidyverse.org](https://tidyverse.org)

#### Everything is found here

.footnote[
https://tidyverse.org/ <br>
https://tidyverse.tidyverse.org/
]



???

The definitive guide and descriptions are found at tidyverse.org.  In fact, the majority of this presentation is borrowed from the official sources.
If you need 1 take home message, it should be "tidyverse.org"

The content on the internet is great.
I should also disclaim, that I am no `tidyverse` expert.
I read (some of) this stuff so you do not have to.

### Shortcuts:
* `h` for help
* `number` + `Enter` go to page
* `b` for "black-out"
* `m` for "mirror"
* `f` to toggle full-screen 
* `c` for clone slides to a new browser window; slides in the two windows will be in sync as you navigate through them
* `p` for presenters mode 




---
class: middle, center

# Thank you



---
class: middle

```{r, comment = " "}
library(tidyverse)
tidyverse_logo()
```


???

Tidyverse is simply a collection of 8 core package with which share common design principles.
Or, more bluntly out: 'all the nice stuff from Hadley Wickham and co. at Rstudio'.

Each package probably warrants it's one presentation.



---
class: center

#### The `tidyverse` dependencies

```{r echo=FALSE, message=FALSE, warning=FALSE}
source("tidyverse_deps.R")$value
```

.left[
Press `ctrl`+`R` or `F5`
]


---
# A common design philsophy


* A shared concept of *tidy data*; the *tidy*verse, not the *messy*verse
     * Programs are for humans to read
     * Embrace functional programming
     * Use and reuse existing data structures
     * Compose simple functions with the pipe

* "There should be one, and preferably only one, obvious way to do it."

    * The **R** core team's philosophy is fundamentally different.

.footnote[
https://principles.tidyverse.org/ <br>
https://github.com/tidyverse/principles/issues <br>
https://tidyverse.tidyverse.org/articles/manifesto.html
]

???

* The traditional way of manipulating data in R is quite different from this
* That obvious way may not be obvious at first.
* R core team says: many ways to do the same thing.
* The grammar of graphics

---
# What is *tidy data*?

> Happy families are all alike; every unhappy family is unhappy in its own way.
> .right[*Leo Tolstoy*]

Data is *tidy* **if**:
1. Each variable forms a column.
2. Each observation forms a row.
3. Each type of observational unit forms a table. (A value must have it own cell)


.footnote[
Wickham, H. (2014). Tidy Data. Journal of Statistical Software, 59(10), 1-23. doi:http://dx.doi.org/10.18637/jss.v059.i10
]

???

* Speaks only toward 'rectangular data' --- there is lots of data that is not naturally rectangular.
* Some 80% of data analysis is used on cleaning and preparing data (Dasu and Johnson 2003)
* You feel that the 1--2 is tautological, but if you think long-vs-wide tables it should be apparent it is not.
* It is sometimes surprisingly difficult to precisely define variables and observations.
    * Rule of thumb: it's easier to describe functional relationships between variables than between rows, 
    * it is easier to make comparisons between groups of observations than between groups of columns. 


---
# What is **not** tidy data?

>* Column headers are values, not variable names.
>* Multiple variables are stored in one column.
>* Variables are stored in both rows and columns.
>* Multiple types of observational units are stored in the same table.
>* A single observational unit is stored in multiple tables.

* Contingency tables / *n*-factor tabulation arrays

.footnote[
Wickham, H. (2014) http://dx.doi.org/10.18637/jss.v059.i10
]

???




---
class: middle

### Tidy?
```{r}
print(AirPassengers) # Monthly Airline Passenger Numbers 1949-1960
```
                




---

## `%>%`

The following are equivalent

- `value %>% f(...)`
- `f(value, ...)`


      Univariate f:    x %>% f          is the same as  f(x)
      Multivariate g:  x %>% g(y, ...)  is the same as  g(x, y, ...)

```{r}
iris %>% 
  subset(Species == "setosa", select = names(.)[-4]) %>%
  head(n=2)
```
```{r, eval=FALSE} 
head(subset(iris, Species=="setosa", select = names(iris)[-4]), n=2)
```


???

Called the pipe operator; performs function composition.
From the package `magrittr`; adopted by `tidyverse`.
Can give some expressive code by daisy chaining pipes.
The dot `.` can be used as placeholder to place the left hand size elsewhere or use
the `value` for other purposes.









